Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-20 manu-draft.Rmd
working on methods, data sources

This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.  We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.  A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Piney Point.  The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage resources from existing partnerships among local resource management institutions.

@@

Discharges from Piney Point did not occur again until 2021.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality measured in NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.  This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge release of Piney Point began in April, 2021 and continued over the following months.

>>>>

This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.  We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.  A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Piney Point.  The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage resources from existing partnerships among local resource management institutions.

@@

Discharges from Piney Point did not occur again until 2021.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.  This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge release of Piney Point began in April, 2021 and continued over the following months. These data were collected through a coordinated effort, facilitated in part by the TBEP.  Monitoring agencies and local partners that collected included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Florida Fish and Wildlife Conservation Commission Fish and Wildlife Research Institute, City of St. Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing and *in situ* measurements of discrete samples), phytoplankton cell counts, benthic sediment samples, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).  Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted but they are not reported here.

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken from surface grabs by boat and processed by the respective laboratories of each participating agency.

## Long-term monitoring data

## Data Analysis
